\thetitle

### 2013-06-19 tfpie/tfpie.tex
End of my edits.

class Examples {
    void test_tree(Tester t) {
        t.checkExpect(new Leaf(7).sum(), 7);
        t.checkExpect(new Node(new Leaf(1),
                               5, 
                               new Node(new Leaf(0), 10, new Leaf(0))).sum(),
                      16);
    }
}
\end{verbatim}
\caption{Binary tree sum in How to Design Classes}
\label{fig:java}
\end{figure}

@@

Thanks to the the preparation in the first half of the course, we can
cover functional OO programming in Java in a just a few lectures.  We
then increase the subset of the language we use to encompass mutation,
loops, and mutable data structures. We present \texttt{ArrayList}s,
followed briefly by arrays. Students use, and then implement, hash
tables as well as other mutable and immutable data structures.
Conventional input and output are treated only very briefly, as we
focus instead of both fundamentals and exercises making use of real
APIs such as hashing functions or Twitter posting.  Finally, while,
for, and for-each loops are presented, following the methodology of
\htdc which connects loops to stylized use of recursive functions with
accumulators, a technique the students now have two semesters of
practice with.

@@

Finally, at the end of the course, we are able to build on the two
major segments to examine less-well-explored topics in object-oriented
programming.  Typically, we cover the basics of implementing OO
programming in a functional language, advanced OO techniques such as
mixins and prototypes, and a new OO language such as Ruby or
JavaScript.  Additionally, we emphasize the ability to embed
functional programming in an OO context, using techniques such as the
command pattern and the visitor patterns.  Again, the key message is
the transferability of concepts across languages.

@@

Teaching programming principles in a functional style has a long
history, with Abelson and Sussman's \emph{Structure and Interpretation
of Computer Programs}~\cite{dvanhorn:sicp} a prominent example.  Our
work follows in the tradition of the \emph{TeachScheme!} project, now
\emph{Program by Design} (PbD), which emphasizes a systematic
approach to program construction.

@@

\subsection{TeachScheme! and ProfessorJ}

The TeachScheme! project initially focused only on the first course,
with the second course typically taught in Java in
institution-specific ways. Subsequently, the pedagogical approach was
extended to Java, but without the tool support and textbook of the
first course.  An example of this approach is described by
\citet{dvanhorn:Bloch2000Scheme}, who presents the experience
integrating these courses at Adelphi.  He reports that ``many of
Java's concepts could be introduced more easily in a second course
than a first''.

With these lessons in mind, the TeachScheme! project set out to apply
the lessons of teaching languages and IDE support to Java, as well as
to present the approach to object-oriented programming in textbook
form. ProfessorJ~\cite{dvanhorn:Gray2003ProfessorJ} is the resulting
system, accompanying the draft textbook \emph{How to Design
  Classes}~\cite{local:htdc}.  In parallel to our course, Northeastern
teaches the remainder of its computer science majors following this
approach.

ProfessorJ and How to Design Classes maintain many of the excellent
ideas of the first course.  In particular, ProfessorJ brings language
levels to Java, in an attempt to smooth the transition for students
from the first course and provide more helpful feedback.  ProfessorJ
is also embedded in the DrRacket IDE, increasing familiarity for the
students and supporting tools such as an interactive read-eval-print
loop.

However, the ``day 1'' transition from the student languages used with
\htdp to \profj is too abrupt and too large.
%
Most significantly, changing languages from the first semester
immediately rather than simply adding a new concept confuses too many
issues for students.  On the first day of a \htdc-based course,
students see object-orientation, a new programming paradigm; Java, a
new language with new syntax, and static types, a crucial but
orthogonal concept.
%
In contrast, our course presents just one of these concepts on the
first day, but  covers all of them by the end of the semester.

@@

Finally, committing to Java on the first day, regardless of the
environment provided to students, has significant limitations.  First,
the syntactic and semantic heaviness of Java is a burden for beginning
students, and discourages interactive and experimental programming.
The very first chapter of \htdc discusses the fixed size of Java integers,
a topic avoided entirely in the first course.
Second, by committing to a particular industrial-strength language, it
closes off possibilities in the curriculum.  Third, it commits
entirely to the new paradigm, making it more difficult for students to
compare the approaches.

@@

Of course, not every curriculum that begins with \htdp transitions to
Java after the first course.  \citet{dvanhorn:Ragde2008Chilling}
describes a second course that includes both more advanced work in
Scheme beyond teaching-oriented languages as well as low-level
programming in C, taught to computer science majors at University of
Waterloo.  Radge's course intentionally does not use student-oriented
languages, although the recently-developed C0 language~\cite{local:c0}
could provide this. Other discussions of functional programming in the
first year~\cite{dvanhorn:Chakravarty2004Risks} do not discuss the
problems of transition.

@@

The course has also provided a vantage point to introduce topics that
will be taken up later in the curriculum.  We present types,
contracts, invariants, and properties of functions, all of which tie
into both the concurrent course on logic and computation, as well as
later classes on formal methods.  The emphasis on
representation-independence and interfaces  both ties into later
classes on software engineering, as well as preparing the ground for
algorithms and data structures courses.  Finally, the use of
interactive and distributed systems connects to later courses on
operating systems and networks.

>>>>

class Examples {
    void test_tree(Tester t) {
        t.checkExpect(new Leaf(7).sum(), 7);
        t.checkExpect(new Node(new Leaf(1),
                               5, 
                               new Node(new Leaf(0), 10, new Leaf(0))).sum(),
                      16);
    }
}
\end{verbatim}
\caption{Binary tree sum in the style of \emph{How to Design Classes}}
\label{fig:java}
\end{figure}

@@

Thanks to the the preparation in the first half of the course, we can
cover functional OO programming in Java in a just a few lectures.  We
then increase the subset of the language we use to encompass mutation,
loops, and mutable data structures and introduce the underlying design
principles. We present \texttt{ArrayList}s, followed briefly by
arrays. Students use, and then implement, hash tables as well as other
mutable and immutable data structures.  Conventional input and output
are treated only very briefly, as we focus instead on both
fundamentals and exercises making use of real APIs such as hashing
functions or Twitter posting.  Finally, \texttt{while}, \texttt{for},
and for-each loops are presented, following the methodology of \htdc
which connects loops to stylized use of recursive functions with
accumulators, a technique the students now have used for two semesters.

@@

Finally, at the end of the course, we are able to build on the two
major segments to examine less-well-explored topics in object-oriented
programming.  Typically, we cover the basics of representing objects
in a functional language, advanced OO techniques such as mixins and
prototypes, and a new OO language such as Ruby or JavaScript.
Additionally, we emphasize the ability to embed functional programming
in an OO context, using techniques such as the command pattern and the
visitor patterns.  Again, the key message is the transferability of
concepts across languages.

@@

Teaching programming principles in a functional style has a long
history, with Abelson and Sussman's \emph{Structure and Interpretation
  of Computer Programs}~\cite{dvanhorn:sicp} being a prominent
example.  Our work follows in the tradition of the \emph{Program by
  Design} (PbD)
project\footnote{\url{http://www.programbydesign.org/}} (previously
known as the \emph{TeachScheme!} project), which emphasizes a
systematic approach to program construction.

@@

\subsection{Program by Design and ProfessorJ}

The Program by Design project initially focused only on the first course,
with the second course typically taught in Java in
institution-specific ways. Subsequently, the pedagogical approach was
extended to Java, but without the tool support and textbook of the
first course.  An example of this approach is described by
\citet{dvanhorn:Bloch2000Scheme}, who presents the experience
integrating these courses at Adelphi.  He reports that ``many of
Java's concepts could be introduced more easily in a second course
than a first.''

With these lessons in mind, the PbD project set out to apply
the lessons of teaching languages and IDE support to Java, as well as
to present the approach to object-oriented programming in textbook
form. ProfessorJ~\cite{dvanhorn:Gray2003ProfessorJ} is the resulting
system, accompanying the draft textbook \emph{How to Design
  Classes}~\cite{local:htdc}.  In parallel to our course, Northeastern
teaches the remainder of its computer science majors following this
approach.

ProfessorJ and \emph{How to Design Classes} maintain many of the ideas
of the first course.  In particular, ProfessorJ brings language levels
to Java, in an attempt to smooth the transition for students from the
first course and provide more helpful feedback.  ProfessorJ is also
embedded in the DrRacket IDE, increasing familiarity for the students
and supporting tools such as an interactive read-eval-print loop.

However, the ``day 1'' transition from the student languages used with
\htdp to \profj is too abrupt and too large.
%
Most significantly, changing languages from the first semester
immediately rather than simply adding a new concept confuses too many
issues for students.  On the first day of a \htdc-based course,
students see object-orientation, a new programming paradigm; Java, a
new language with new syntax, and a static type system, a crucial but
orthogonal concept.
%
In contrast, our course presents just one of these concepts on the
first day, but  covers all of them by the end of the semester.

@@

Committing to Java on the first day, regardless of the environment
provided to students, has significant limitations.  First, the
syntactic and semantic heaviness of Java is a burden for beginning
students, and discourages interactive and experimental programming.
The very first chapter of \htdc discusses the fixed size of Java
integers, a topic avoided entirely in the first course.  Second, by
committing to a particular industrial-strength language, it closes off
possibilities in the curriculum.  Third, it commits entirely to the
new paradigm, making it more difficult for students to compare the
approaches.

@@

Not every curriculum that begins with \htdp transitions to Java after
the first course.  \citet{dvanhorn:Ragde2008Chilling} describes a
second course that includes both more advanced work in Scheme beyond
teaching-oriented languages as well as low-level programming in C,
taught to computer science majors at University of Waterloo.  Radge's
course intentionally does not use student-oriented languages, although
the recently-developed C0 language~\cite{local:c0} could provide such
a language. Other discussions of functional programming in the first
year~\cite{dvanhorn:Chakravarty2004Risks} do not discuss the problems
of transition.

@@

The course has provided a vantage point to introduce topics that will
be taken up later in the curriculum.  We present types, contracts,
invariants, and properties of functions, all of which tie into both
the concurrent course on logic and computation, as well as later
classes on formal methods.  The emphasis on
representation-independence and interfaces both tie into later classes
on software engineering, as well as preparing students for algorithms
and data structures courses.  Finally, the use of interactive and
distributed systems connects to later courses on operating systems and
networks.
